---
title: "Arab_Barometer_Analysis"
author: "Yuna Baek"
date: "2025-03-14"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Libraries
```{r}
library(haven) 
library(readr)
library(dplyr)
library(tidyr)
library(corrplot)
library(ggplot2)
library(ggthemr)
library(gplots)
library(modelsummary)
```

```{r}
path.to.data <- "/Users/yunabaek/Desktop/Harris/Year 2/2025 Winter/2. Policy Lab/Policy-Lab/Data/ArabBarometer_WaveVIII_English_v1.csv"
arab_df <- read_csv(path.to.data)
```


```{r}
# Function to generate correlation matrix heatmap
create_heatmap <- function(cor_matrix, title) {
  corrplot(cor_matrix,
           method = "color",
           type = "upper",
           tl.col = "black",
           tl.srt = 45,
           addCoef.col = "black",
           number.cex = 0.7,
           title = title,
           mar = c(0,0,2,0))
}
```

### Data wrangling ### 
```{r}
# Recode 98, 99 as NA
arab_df <- as.data.frame(lapply(arab_df, function(x) {
  x[x %in% c(98, 99)] <- NA
  return(x)
}))
```

### Dependent Variable: Q551A ###
```{r}
# Dummy variable for Q551A
for (i in 1:6) {
  arab_df[[paste0("Q551A_", i)]] <- ifelse(is.na(arab_df$Q551A), NA, 
                                           ifelse(arab_df$Q551A == i, 1, 0))
}
Q551A_vars <- grep("^Q551A_", names(arab_df), value = TRUE)
```

Demographics variables
- Gender - Q1002: 1=male
- Education - Q1003: 1=no formal educ, 2=elementary, 3=preparatory/basic, 4=secondary, 5=mid-level diploma, 6=BA, 7=MA&above
- Urban/Rural - Q13: 1=urban, 2=rural, 3=camp
```{r}
arab_df$female <- ifelse(is.na(arab_df$Q1002), NA, ifelse(arab_df$Q1002 == 1, 0, 1)) # If male, 0
arab_df$secondary <- ifelse(is.na(arab_df$Q1003), NA, ifelse(arab_df$Q1003 >= 4, 1, 0)) # If higher than or equal to secondary, 1
arab_df$rural <- ifelse(is.na(arab_df$Q13), NA, ifelse(arab_df$Q13 == 2, 1, 0)) # If 1, rural
```

Balance test of demographics against Q551A
```{r}
balance_test <- function(df, group_var, outcome_var) {
  tmp <- df %>%
    filter(!is.na({{group_var}}), !is.na({{outcome_var}})) %>%
    group_by({{group_var}}) %>%
    summarise(
      Count = sum({{outcome_var}}, na.rm = TRUE),  # sum of 1's
      N     = n(),                                 # total rows
      .groups = "drop"
    ) %>%
    mutate(Prop = Count / N)
  
  out <- tmp %>%
    summarise(
      Group0_Prop  = ifelse(any({{group_var}} == 0), Prop[{{group_var}} == 0], NA),
      Group1_Prop  = ifelse(any({{group_var}} == 1), Prop[{{group_var}} == 1], NA),
      Group0_Count = ifelse(any({{group_var}} == 0), Count[{{group_var}} == 0], NA),
      Group1_Count = ifelse(any({{group_var}} == 1), Count[{{group_var}} == 1], NA),
      Group0_Total = ifelse(any({{group_var}} == 0), N[{{group_var}} == 0], NA),
      Group1_Total = ifelse(any({{group_var}} == 1), N[{{group_var}} == 1], NA)
    )
  
  # Two‐proportion z‐test p-value
  out <- out %>%
    mutate(
      p_value = if (!is.na(Group0_Count) && !is.na(Group1_Count) &&
                    Group0_Count > 0 && Group1_Count > 0) {
        prop.test(
          x = c(Group0_Count, Group1_Count),
          n = c(Group0_Total, Group1_Total),
          correct = FALSE
        )$p.value
      } else {
        NA
      }
    )
  
  out
}

results <- lapply(1:6, function(i) {
  var_name <- paste0("Q551A_", i)
  
  # -- Gender
  tmp_gender <- balance_test(arab_df, female, !!sym(var_name)) %>%
    rename(
      Male_Prop   = Group0_Prop,
      Female_Prop = Group1_Prop,
      p_gender    = p_value
    )
  
  # -- Setting
  tmp_rural <- balance_test(arab_df, rural, !!sym(var_name)) %>%
    rename(
      Urban_Prop = Group0_Prop,
      Rural_Prop = Group1_Prop,
      p_rural    = p_value
    )
  
  # -- Education
  tmp_edu <- balance_test(arab_df, secondary, !!sym(var_name)) %>%
    rename(
      Low_Prop  = Group0_Prop,
      High_Prop = Group1_Prop,
      p_edu     = p_value
    )

  data.frame(
    Q551A_i = i,
    Male = tmp_gender$Male_Prop,
    Female = tmp_gender$Female_Prop,
    p_value_gender = tmp_gender$p_gender,
    
    Urban = tmp_rural$Urban_Prop,
    Rural = tmp_rural$Rural_Prop,
    p_value_rural = tmp_rural$p_rural,
    
    Low = tmp_edu$Low_Prop,
    High = tmp_edu$High_Prop,
    p_value_edu = tmp_edu$p_edu
  )
}) %>%
  bind_rows()
results <- results %>%
  mutate(
    Male      = round(Male * 100, 1),
    Female    = round(Female * 100, 1),
    Urban     = round(Urban * 100, 1),
    Rural     = round(Rural * 100, 1),
    Low       = round(Low * 100, 1),
    High      = round(High * 100, 1),
      Q551A_label = case_when(
      Q551A_i == 1 ~ "Equality under the law",
      Q551A_i == 2 ~ "Absence of corruption",
      Q551A_i == 3 ~ "Basic necessities available",
      Q551A_i == 4 ~ "Free election choice",
      Q551A_i == 5 ~ "Safety from danger",
      Q551A_i == 6 ~ "Civil rights guaranteed",
      TRUE ~ NA_character_
    )
  ) %>%
  select(Q551A_i, Q551A_label, everything())

print(results)
```

Financial Stability - Converting to binary variables
```{r}
# income stability = 1 if "able to save / no difficulties", 0 if "face some / significant difficulties"
arab_df$income_stability <- ifelse(is.na(arab_df$Q1016), NA, 
                                   ifelse(arab_df$Q1016 %in% c(1, 2), 1, 0))

arab_df$has_savings <- ifelse(is.na(arab_df$Q116), NA, 
                              ifelse(arab_df$Q116 == 1, 1, 0))
arab_df$bank_account <- ifelse(is.na(arab_df$Q1025), NA, 
                               ifelse(arab_df$Q1025 == 1, 1, 0))

# 1 if own current place of residence
arab_df$hh_own <- ifelse(is.na(arab_df$Q1001G), NA, 
                         ifelse(arab_df$Q1001G == 1, 0, 1))

# if own current house & has no mortgage = 1
arab_df$home_own_no_mortgage <- ifelse(is.na(arab_df$Q1001H), NA, 
                                       ifelse(arab_df$Q1001H == 1, 0, 1))

arab_df$above_median_income <- ifelse(is.na(arab_df$Q1015A), NA, 
                                      ifelse(arab_df$Q1015A == 2, 1, 0))

stable_jobs <- c(1, 2, 4, 6)  # Salaried, stable hourly income, rent/financial products
arab_df$employment_stability <- ifelse(is.na(arab_df$Q1021), NA, 
                                    ifelse(arab_df$Q1021 %in% stable_jobs, 1, 0))

pca_vars <- c("income_stability", "has_savings", "bank_account", "hh_own", 
              "home_own_no_mortgage", "above_median_income","employment_stability")
```

Histogram of financial variables distribution
```{r}
library(gridExtra) 

# Function to create bar plots including NA values
plot_bar <- function(variable, title) {
  variable <- factor(variable, levels = unique(c(variable, NA)), exclude = NULL)
  data_tbl <- as.data.frame(table(variable, useNA = "always"))
  data_tbl$variable <- as.character(data_tbl$variable)
  data_tbl$variable[is.na(data_tbl$variable)] <- "NA"
  # Create bar plot
  ggplot(data_tbl, aes(x = variable, y = Freq)) +
    geom_bar(stat = "identity", fill = "skyblue", alpha = 0.7) +
    labs(title = title, x = "Binary Response", y = "Frequency") +
    theme_minimal() +
    theme(axis.text.x = element_text(hjust = 1))
}

# Generate bar plots
p1 <- plot_bar(arab_df$income_stability, "Income Stability")
p2 <- plot_bar(arab_df$has_savings, "Have Savings")
p3 <- plot_bar(arab_df$bank_account, "Own a Bank Account")
p4 <- plot_bar(arab_df$hh_own, "Own Current Place of Residence")
p5 <- plot_bar(arab_df$home_own_no_mortgage, "Own Home & No Mortgage")
p6 <- plot_bar(arab_df$above_median_income, "Less or More Than Median Income")
p7 <- plot_bar(arab_df$employment_stability, "Employment Stability")

# Print plots
grid.arrange(p1, p2, p3, p4, p5, p6, p7, ncol = 2)
```

Basic necessities: Reversing, correlation 
```{r}
# Basic necessities
# Reverse coding for Basic Necessities variables (Higher = Worse Condition)
arab_df <- arab_df %>% 
  mutate(
    # Q130: Food Insecurity Indicator (Binary: 0 = Any Coping Mechanism Used, 1 = Not food insecure)
    Q130_r = if_else(Q130_2 == 1 | Q130_3 == 1 | Q130_4 == 1 | Q130_5 == 1, 0, 1, missing = NA_real_)
    )

# Replace 97 with NA
arab_df[c("Q540B_3", "Q274A", "Q274B")] <- lapply(arab_df[c("Q540B_3", "Q274A", "Q274B")], function(x) {
  x[x %in% c(97)] <- NA
  return(x)
})

necessities_vars <- c("Q129A_1", "Q129A_2", "Q130_r", "Q112_2",
                      "Q540B_3", "Q274A", "Q274B")

necessities_cors <- cor(arab_df[necessities_vars],
                       use = "pairwise.complete.obs",
                       method = "spearman")

necessities_vars <- c("Q129A_1", "Q129A_2", "Q130_r", "Q112_2",
                      "Q540B_3", "Q274A", "Q274B")


create_heatmap(necessities_cors, "Correlations: Basic Necessities")

```


### Creating SES Indices ###

# Food security index
```{r}
# Scaling [0, 1]
library(caret)

# Food security & access to services
# Selecting indices based on correlation results
food_security_vars <- c("Q129A_1", "Q129A_2", "Q130_r", "Q112_2")
access_vars <- c("Q540B_3", "Q274A", "Q274B")
columns_to_scale <- c(food_security_vars, access_vars)

# Apply scaling to selected columns
scale_model <- preProcess(arab_df[, columns_to_scale], method = c("range"))
scaled_data <- predict(scale_model, arab_df[, columns_to_scale])

# Add new scaled columns to arab_df with names Q129A_1_scaled
arab_df[, paste0(columns_to_scale, "_scaled")] <- scaled_data

# Summary index
arab_df$food_security_index <- rowMeans(arab_df[paste0(food_security_vars, "_scaled")], na.rm = TRUE)
arab_df$access_index <- rowMeans(arab_df[paste0(access_vars, "_scaled")], na.rm = TRUE)
```


Correlation matrix of SES variables
```{r}
financial_cors <- cor(arab_df[c(pca_vars, "food_security_index", "access_index", "secondary")],
                       use = "pairwise.complete.obs",
                       method = "spearman")

# Create the heatmap for the Absence of Corruption correlations
create_heatmap(financial_cors, "Correlations: SES Variables")
```

Summary Index 1: financial stability index
```{r}
# Mean index using Q1001G, Q1025, Q1016; binary vars 
arab_df$financial_stability_index <-  rowSums(
  arab_df[, c("income_stability", "has_savings", "bank_account")], 
  na.rm = TRUE
)
```

Quartile for binary conversion
```{r}
# Calculate quartile cutoffs
financial_quartiles <- quantile(arab_df$financial_stability_index, probs = c(0.25, 0.75), na.rm = TRUE)

# Assign quartile-based categories (Low, Middle, High)
arab_df$financial_quartiles <- cut(arab_df$financial_stability_index, 
                            breaks = c(-Inf, financial_quartiles[1], financial_quartiles[2], Inf), 
                            labels = c("Low", "Middle", "High"),
                            include.lowest = TRUE)

# Check distribution
table(arab_df$financial_quartiles, useNA = "always")

# Create binary financial stability variable: High = Top 25%, Low = Bottom 75%
arab_df$financial_binary <- ifelse(is.na(arab_df$financial_stability_index), NA,
                             ifelse(arab_df$financial_stability_index >= financial_quartiles[2], 1, 0))

# Check distribution
table(arab_df$financial_binary, useNA = "always")
```

Summary Index 2: economic stability index based on the high correlation coefficients
```{r}
# Mean index using two binary vars & food security index
arab_df$econ_stability_index <-  rowSums(
  arab_df[, c("income_stability", "above_median_income", "food_security_index")], 
  na.rm = TRUE
)
```

Alternative option: using median or quartile cutoffs to convert indices to binary indices
```{r}
# Calculate quartile cutoffs
econ_quartiles <- quantile(arab_df$econ_stability_index, probs = c(0.25, 0.75), na.rm = TRUE)

# Assign quartile-based categories (Low, Middle, High)
arab_df$econ_quartiles <- cut(arab_df$econ_stability_index, 
                            breaks = c(-Inf, econ_quartiles[1], econ_quartiles[2], Inf), 
                            labels = c("Low", "Middle", "High"),
                            include.lowest = TRUE)

# Check distribution
table(arab_df$econ_quartiles, useNA = "always")

# Create binary financial stability variable: High = Top 25%, Low = Bottom 75%
arab_df$econ_binary <- ifelse(is.na(arab_df$econ_stability_index), NA,
                             ifelse(arab_df$econ_stability_index >= econ_quartiles[2], 1, 0))

# Check distribution
table(arab_df$econ_binary, useNA = "always")
```


# Regression Analysis
## Logit Model Summary Table with Q551A_i as dependent variables
Function: run_glm_models
```{r}
# Define dependent and demographic variables
dep_vars <- Q551A_vars
dem_vars <- c("female", "secondary", "rural")

# Define custom labels for dependent variables
dep_labels <- c(
  "Q551A_1" = "Equality under the law",
  "Q551A_2" = "Absence of corruption",
  "Q551A_3" = "Basic necessities",
  "Q551A_4" = "Freely choose political leaders",
  "Q551A_5" = "Physical safety",
  "Q551A_6" = "Basic civil rights"
)

# Function inputs: independent vars, disc_labels, table title
run_glm_models <- function(indep_vars, disc_labels, title) {
  models_list <- list()

  for (dep in dep_vars) {
    # Regression input formula
    fml <- as.formula(paste(dep, "~", paste(indep_vars, collapse = " + ")))

    # Subset to complete cases
    subset_data <- arab_df[complete.cases(arab_df[, c(dep, indep_vars)]), ]

    # Check that there is data and that the dependent variable has at least 2 unique values
    if (nrow(subset_data) == 0 || length(unique(subset_data[[dep]])) < 2) {
      models_list[[dep_labels[dep]]] <- NA
    } else {
      model <- glm(fml, data = subset_data, family = binomial, na.action = na.omit)
      models_list[[dep_labels[dep]]] <- model # mar5 5:21pm
      DescTools::PseudoR2(model, which = "all")
    }
  }

# Generate the regression table
  modelsummary(models_list,
               stars = TRUE,
               title = title,
               coef_map = disc_labels)
}
```


GLM model: Demographic variables on Q551A
```{r}
dem_labels <- c(
  "female" = "Female",
  "secondary" = "Secondary Education",
  "rural" = "Rural"
)
run_glm_models(dem_vars, dem_labels, "Logit: Q551A and Demographic Variables")
```

Food Security Index
```{r}
indep_vars_1 <- c("food_security_index")
food_security <- run_glm_models(indep_vars_1, disc_labels_1, "Logit: SES Indices")
```

```{r}
# Model with demographic variables
indep_vars_1 <- c("food_security_index")
disc_labels_1 <- c(
  "food_security_index" = "Food Security Index"
)
run_glm_models(indep_vars_1, disc_labels_1, "Logit: Indices without Demographic Controls")

# Model without demographic variables
indep_vars_2 <- c(indep_vars_1, dem_vars)
disc_labels_2 <- c(disc_labels_1, dem_labels)
run_glm_models(indep_vars_2, disc_labels_2, "Logit: Indices with Demographic Controls")

```



## Interaction Term Logit Model Summary Table with Q551A_i as dependent variables
glm_int_models
interaction terms = demographic variables (demo_vars)
```{r}
# inputs: independent vars, demographic vars (interaction), disc_labels, title 
glm_int_models <- function(indep_vars, demo_vars, disc_labels, title) {
  models_list <- list()
  
  for (dep in dep_vars) {
    # Create main effects: combine independent and demographic variables
    main_effects <- paste(c(indep_vars, demo_vars), collapse = " + ")
    
    # Create interaction terms for each independent variable with each demographic variable
    interaction_terms <- c()
    for (indep in indep_vars) {
      for (demo in demo_vars) {
        interaction_terms <- c(interaction_terms, paste0(indep, ":", demo))
      }
    }
    
    # Combine main effects and interaction terms into a full regression formula
    fml_str <- paste(dep, "~", main_effects, "+", paste(interaction_terms, collapse = " + "))
    fml <- as.formula(fml_str)
    
    # Subset data to complete cases for the dependent variable, independent variables, and demo variables
    subset_data <- arab_df[complete.cases(arab_df[, c(dep, indep_vars, demo_vars)]), ]
    
    # Check for sufficient data and variation in the dependent variable
    if (nrow(subset_data) == 0 || length(unique(subset_data[[dep]])) < 2) {
      models_list[[dep_labels[dep]]] <- NA
    } else {
      model <- glm(fml, data = subset_data, family = binomial, na.action = na.omit)
      models_list[[dep_labels[dep]]] <- model
    }
  }
  
  # Generate the regression table with the specified coefficient labels
  modelsummary(models_list,
               stars = TRUE,
               title = title,
               coef_map = disc_labels)
}
```


```{r}
### Regression with interaction terms with demographic variables

# Define independent and demographic variables
indep_vars <- c("food_security_index")
demo_vars <- c("female", "secondary", "rural")

# Define labels for the main effects
disc_labels <- c(
  "food_security_index" = "Food Security Index",
  "female" = "Female",
  "secondary" = "Secondary Education",
  "rural" = "Rural"
)

# Create dynamic labels for each interaction term (e.g., "Water & Electricity Access Index × Female")
interaction_labels <- c()
for (indep in indep_vars) {
  for (demo in demo_vars) {
    interaction_term <- paste0(indep, ":", demo)
    interaction_labels[interaction_term] <- paste0(disc_labels[indep], " × ", disc_labels[demo])
  }
}

# Add the interaction labels to the coefficient labels list
disc_labels <- c(disc_labels, interaction_labels)

# Call the function to run the models with interaction terms
glm_int_models(indep_vars, demo_vars, disc_labels, "Logit: Q551A and Demographic Variables with Interactions")

```



